Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-03-10 manu_draft.Rmd
starting revisions from PDV, no more mixed-meta language

Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.  A key requirement is complete propagation of uncertainty through the analysis.  However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods. Here, we propose a novel application of generalized additive models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.  GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.  These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.  Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.  We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.  We then extract seasonal averages for an *a priori* period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.  Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluate changes in the seasonal averages between years with full propagation of uncertainty in the estimates.  Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.

@@

Finally, trend results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.  As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.  Moreover, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.  Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.  Trend estimates from mixed-meta regression applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages. Select examples were used where differences were pronounced to illustrate false positive or negatives that may occur with alternative methods. This analysis was then applied to all stations.

@@

Slope estimates that were used to determine significance of a seasonal trend across years from the mixed-meta analysis were not back-transformed.  While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.  As such, all results reporting slope estimates as rates of change in chl-a per year were reported at the scale of the model in log$_{10}$-space.

@@

Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.  Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations. Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased at stations 30 and 32 from 2010 to 2019. For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.

@@

Results showing trend estimates from mixed-meta regression on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from observed data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates (row 2). Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.  Only the mixed-meta regression results show a trend in this example (based on $p < 0.05$).  The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.  Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.  Unlike the first example, the top two figures show trends, whereas the bottom plot for the mixed-meta regression analysis does not show a trend.  In both cases, only the mixed-meta regression results provide accurate trend estimates because of full propagation of uncertainty across methods.

Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to June period.  Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.  Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.

@@

Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons demonstrate a common approach where raw observations may be aggregated prior to use of more conventional trend analysis methods.  Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the mixed-meta regression results.  Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, but the mixed-meta results did not.  The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).

Variability of sampling effort among years is a common characteristic of environmental monitoring data, which further increases the potential for incorrect conclusions when existing methods have been used. Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months [e.g., summer only, @Fouquet12].  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].  As a result, trend assessments from mixed-meta regression are more accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.

@@

Additional work could be conducted to further strengthen the validity of conclusions based on trends from mixed-meta regression analyses applied to the GAM seasonal averages.  We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends.  Although there are undoubtedly many scenarios where years of interest can be chosen objectively by the needs of an analysis (e.g., regulatory compliance periods, time since management intervention), a more general question of "when" changes occur independent of user decisions is also important to address.  For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?  There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.  Assessing water quality changes beyond an evaluation of seasonal averages could also be possible with our approach, such as assessing changes in the timing or magnitude of a seasonal peak across years.

@@

png('figs/trnddat.png', height = 6, width = 11, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates in ten year blocks from mixed-meta analyses using results of GAM predictions for station 34.  Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.  The trend lines estimate the rate of change of chl-a per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title. ns: not significant at $\\alpha$ = 0.05, * p < 0.05, ** p < 0.005"}
knitr::include_graphics('figs/trnddat.png')
```

@@

png('figs/winchg.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r winchg, fig.cap='Estimates of log$_{10}$ chl-a change per year (+/- 95\\% confidence interval) from applying the mixed-meta analysis across the seasonal averages for each station.  Stations are arranged top to bottom from north to south.  Plots in (a) show estimates for seasonal averages from January to June and plots in (b) show estimates for seasonal averages from July to December. Results are from a ten-year, centered moving window where each point shows a linear trend estimate from five years prior to after each year. Estimates prior to 1996 and after 2014 are not available because of an incomplete ten year record for estimating the trend. Estimates in a year that are significant are shown in red.'}
knitr::include_graphics('figs/winchg.png')
```

@@

  p3 <- ggplot(data = avgseas, aes(x = yr, y = bt_avg)) +
    geom_point(colour = 'deepskyblue3') +
    geom_errorbar(aes(ymin = bt_lwr, ymax = bt_upr), colour = 'deepskyblue3') +
    geom_ribbon(data = toplo3, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'pink', alpha = 0.4) +
    geom_line(data = toplo3, color = 'pink') +
    theme_minimal(base_family = 'serif') +
    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
    theme(
      axis.title.x = element_blank()
    ) +
    labs(
      subtitle = paste('Mixed-meta regression on GAM estimates', pval3),
      y = NULL
    )

@@

p <- (p1 |  p2) + plot_layout(ncol = 2)
png('figs/trndcmpex.png', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at  (a) station 36 and (b) station 22. The first row shows OLS (ordinary least squares) regression applied to annual averages of chl-a from the observed data, the second row shows OLS regression applied to annual averages of chl-a from the GAM, and the third row shows mixed-meta regression applied to the annual averages of chl-a from the GAM.  Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
knitr::include_graphics('figs/trndcmpex.png')
```

@@

png('figs/trndcmp.png', height = 4.5, width = 7.5, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station. The "OLS observed" trend model is based on an ordinary least squares (OLS) regression fit to the annual averages of chl-a from the observed data, the "OLS GAM" trend model is based on an OLS regression fit to the annual averages of chl-a from the GAM model, and the "Mixed-meta GAM" trend model is based on a mixed-meta regression fit to the annual averages of chl-a from the GAM model.  Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting trend significance.'}
knitr::include_graphics('figs/trndcmp.png')
```

>>>>

Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.  A key requirement is complete propagation of uncertainty through the analysis.  However, this is difficult when there are mismatches between time scales of monitoring data and trends of interest. Here, we propose a novel application of generalized additive models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.  GAMs can model a response variable as the additive sum of one or more smoothing splines fit to predictor variables.  These models have shown promise in water quality trend analysis to separate long-term (i.e., annual or decadal) trends from seasonal variation.  Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-effects meta-analysis regression to quantify inter-annual trends that account for full propagation of error across methods.  We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.  We then extract seasonal averages and their standard errors for an *a priori* time period within each year from the GAM results.  Finally, we demonstrate how across-year trends in seasonal averages can be modeled with mixed-effects meta-analysis regression that propagates uncertainties from the GAM fits to the across-year analysis. Overall, this approach leverages GAMs to smooth data with missing observations or varying sample effort across years to estimate seasonal averages and meta-analysis to estimate trends across years.

@@

Finally, trend results from the meta-analysis regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.  As stated above, meta-analysis allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.  Moreover, the meta-analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.  Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.  Trend estimates from meta-analysis applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages. Select examples were used where differences were pronounced to illustrate false positive or negatives that may occur with alternative methods. This analysis was then applied to all stations.

@@

Slope estimates that were used to determine significance of a seasonal trend across years from the meta-analysis were not back-transformed.  While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.  As such, all results reporting slope estimates as rates of change in chl-a per year were reported at the scale of the model in log$_{10}$-space.

@@

Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.  Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations. Mixed-effects meta-analysis regressions applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased at stations 30 and 32 from 2010 to 2019. For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.

@@

Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from observed data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates (row 2). Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.  Only the meta-analysis regression results show a trend in this example (based on $p < 0.05$).  The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.  Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.  Unlike the first example, the top two figures show trends, whereas the bottom plot for the meta-analysis regression does not show a trend.  In both cases, only the meta-analysis results provide accurate trend estimates because of full propagation of uncertainty across methods.

Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and meta-analysis methods, with much larger slopes observed especially at the more southern stations in the January to June period.  Slope estimates from the OLS models applied to the averages from the GAMs as compared to the meta-analysis results were more similar, excluding some of the slope estimates for the southern stations.  Differences in significance of trends between the OLS models applied to the GAM averages and the meta-analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.

@@

Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  These comparisons demonstrate a common approach where raw observations may be aggregated prior to use of more conventional trend analysis methods.  Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the meta-analysis regression results.  Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, but the meta-analysis results did not.  The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).

Variability of sampling effort among years is a common characteristic of environmental monitoring data, which further increases the potential for incorrect conclusions when existing methods have been used. Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months [e.g., summer only, @Fouquet12].  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages. More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-effects meta-analyses [@Sera19].  As a result, trend assessments from meta-analyses are more accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.

@@

Additional work could be conducted to further strengthen the validity of conclusions based on trends from meta-analysis regression applied to the GAM seasonal averages.  We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends.  Although there are undoubtedly many scenarios where years of interest can be chosen objectively by the needs of an analysis (e.g., regulatory compliance periods, time since management intervention), a more general question of "when" changes occur independent of user decisions is also important to address.  For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?  There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.  Assessing water quality changes beyond an evaluation of seasonal averages could also be possible with our approach, such as assessing changes in the timing or magnitude of a seasonal peak across years.

@@

png('figs/trnddat.png', height = 6, width = 11, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates in ten year blocks from meta-analyses using results of GAM predictions for station 34.  Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.  The trend lines estimate the rate of change of chl-a per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title. ns: not significant at $\\alpha$ = 0.05, * p < 0.05, ** p < 0.005"}
knitr::include_graphics('figs/trnddat.png')
```

@@

png('figs/winchg.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r winchg, fig.cap='Estimates of log$_{10}$ chl-a change per year (+/- 95\\% confidence interval) from applying the meta-analysis across the seasonal averages for each station.  Stations are arranged top to bottom from north to south.  Plots in (a) show estimates for seasonal averages from January to June and plots in (b) show estimates for seasonal averages from July to December. Results are from a ten-year, centered moving window where each point shows a linear trend estimate from five years prior to after each year. Estimates prior to 1996 and after 2014 are not available because of an incomplete ten year record for estimating the trend. Estimates in a year that are significant are shown in red.'}
knitr::include_graphics('figs/winchg.png')
```

@@

  p3 <- ggplot(data = avgseas, aes(x = yr, y = bt_avg)) +
    geom_point(colour = 'deepskyblue3') +
    geom_errorbar(aes(ymin = bt_lwr, ymax = bt_upr), colour = 'deepskyblue3') +
    geom_ribbon(data = toplo3, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'pink', alpha = 0.4) +
    geom_line(data = toplo3, color = 'pink') +
    theme_minimal(base_family = 'serif') +
    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
    theme(
      axis.title.x = element_blank()
    ) +
    labs(
      subtitle = paste('Meta-analysis regression on GAM estimates', pval3),
      y = NULL
    )

@@

p <- (p1 |  p2) + plot_layout(ncol = 2)
png('figs/trndcmpex.png', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at  (a) station 36 and (b) station 22. The first row shows OLS (ordinary least squares) regression applied to annual averages of chl-a from the observed data, the second row shows OLS regression applied to annual averages of chl-a from the GAM, and the third row shows meta-analysis regression applied to the annual averages of chl-a from the GAM.  Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
knitr::include_graphics('figs/trndcmpex.png')
```

@@

png('figs/trndcmp.png', height = 4.5, width = 7.5, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station. The "OLS observed" trend model is based on an ordinary least squares (OLS) regression fit to the annual averages of chl-a from the observed data, the "OLS GAM" trend model is based on an OLS regression fit to the annual averages of chl-a from the GAM model, and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the annual averages of chl-a from the GAM model.  Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting trend significance.'}
knitr::include_graphics('figs/trndcmp.png')
```
